Generation of TypeScript Declaration Files from JavaScript Code

### 2021-05-20 mplr-2021/main.tex
Add example for module-class

% I think the reader gets the impression from the text that obtaining the code examples
% is a "main feature" of our tool, where in fact it is only a way of easily obtaining
% runtime information to execute our tool. We could've followed the approach of obtaining
% the runtime information from the tests.
% This can be useful also to divert the attention from how many repositories had README files, etc.
% After using the code examples, we also found out that they provided a useful approximation
% not only on the input types but also on how to use the library (i.e. we don't analyze the structure of the library).
% This is a useful insight which can be used later to infer the correct types.
%
% Should we change this to 'we rely on runtime information'. 'We used the examples
% provided by the developer as part of the documentation of a library as a first
% pragmatic way of obtaining runtime information'.
We aim to improve on this situation by providing a tool that generates
TypeScript declaration files automatically. To do so, we rely on the examples provided
by the developer as part of the documentation of a library. If these
examples are sufficiently rich, then our toolchain can generate
high-quality declaration files for the library. This way we take
advantage of 
best practices in the dynamic languages community which favors
examples and tests over writing out type signatures.

@@

Our tool \texttt{dts-generate} is a first step to explore the possibilities for
generating useful declaration files from code
examples.
% Should we change it to 'explore the possibilities for
% generating useful declaration files from runtime information'.

@@

As \texttt{dts-generate} is based on run-time information, code
fragments are needed that import and use the JavaScript library to
obtain run-time information from running the instrumented library
code.

@@

% I think we should rephrase this, specially now that we know
% it might be 'doable'.

@@

@@

% This is no longer true. I'll add one result to the report.
We were unable to  generate a declaration file based on the
\textbf{module-class} template. All modules corresponding to this template exhibited at least one of
the unimplemented TypeScript features.

>>>>

We aim to improve on this situation by providing a tool that generates
TypeScript declaration files automatically. To do so, we rely on we rely on runtime information
gathered from the examples provided
by the developer as part of the documentation of a library. If these
examples are sufficiently rich, then our toolchain can generate
high-quality declaration files for the library. This way we take
advantage of 
best practices in the dynamic languages community which favors
examples and tests over writing out type signatures.

@@

Our tool \texttt{dts-generate} is a first step to explore the possibilities for
generating useful declaration files from runtime information.

@@

As \texttt{dts-generate} is based on run-time information, 
exemplary code fragments that execute the JavaScript library are
needed to obtain 
significant run-time information from running the instrumented library
code.

@@

@@

\begin{figure*}[tp]
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=TypeScript,numbers=none]
export = SteamID;
declare class SteamID {
    // ...
    isValid(): boolean;
    isGroupChat(): boolean;
    isLobby(): boolean;
    getSteam2RenderedID(newerFormat?: boolean): string;
    // ...
}
declare namespace SteamID {
    export function fromIndividualAccountID(accountid: string | number): SteamID;
}  
    \end{lstlisting}
    \caption{steamid/index.d.ts - Generated (simplified)}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=TypeScript,numbers=none]
// ...
type getSteam2RenderedID = (newerFormat?: boolean) => string;

declare class SteamID {
    // ...
    isValid(): boolean;
    // ...
    isGroupChat(): boolean;
    // ...
    isLobby(): boolean;
    getSteam2RenderedID: getSteam2RenderedID;
    // ...
}

declare namespace SteamID {
    // ...
    function fromIndividualAccountID(accountid: number | string): SteamID;
}

export = SteamID;
    \end{lstlisting}
    \caption{steamid/index.d.ts - DT (simplified)}
  \end{subfigure}

  \caption{Results for \textbf{module-class} template for module \texttt{steamid}}
  \label{fig:experiments-results-module-steamid}
\end{figure*}

@@

\figref{fig:experiments-results-module-steamid} exposes a simplified example for the
\textbf{module-class} template where we highlight the most important features. Class methods are
correctly generated. For the \texttt{getSteam2RenderedID} method we correctly generate a signature
that matches the type alias chosen by the author of the library. Finally, the method \texttt{fromIndividualAccountID}
is correctly located in the namespace section.
